---
title: "Why magnitude-dependent limits of agreement? A CT reader-study walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why magnitude-dependent limits of agreement? A CT reader-study walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5, dpi = 96)
```

This vignette works through one dataset from start to finish. The aim is
not only to show which functions to call, but to show *why* limits of
agreement that follow the magnitude of the measurement are needed, using a
setting where the consequence of getting them wrong is concrete: the RECIST
assessment of tumor response.

## 1. The data

`recist_readings` is a synthetic dataset that reproduces the design of a
published multi-reader CT study (Yoon, Yoon and Hahn, 2019): 249 target
lesions, each read by six radiologists in two replicate sessions, with the
longest diameter recorded in whole millimetres. The data are simulated, so
that they can be shipped with the package, but their error structure is
the one observed in that study: measurement error grows with lesion size,
more slowly than in proportion.

```{r data}
library(flexloa)
data(recist_readings)
head(recist_readings)
summary(recist_readings$diameter)
```

In routine practice a lesion is read once by one radiologist at baseline
and once, often by a different radiologist, at follow-up. The relevant
question is therefore: *how much can a single reading change when the
reader changes?* `loa_pairs()` forms the corresponding pairs, one reading
by one reader against one reading by another reader of the same lesion in
the same session (`replicates = "matched"`, the default).

```{r pairs}
p_inter <- loa_pairs(recist_readings$diameter, recist_readings$lesion,
                     recist_readings$reader, recist_readings$session,
                     type = "inter")
p_inter
```

## 2. The problem: conventional limits on these data

Start where everyone starts, with the conventional Bland-Altman analysis on
the original scale.

```{r constant}
f_id <- loa_fit(p_inter, transform = "identity")
f_id
```

The limits are about ±8 mm for every lesion. Now look at the plot.

```{r constant-plot, fig.cap = "Conventional (constant) limits on the inter-reader pairs."}
plot(f_id, xlab = "Mean of two readers (mm)", ylab = "Difference between readers (mm)")
```

The differences fan out: among lesions of 10-20 mm the readers rarely
disagree by more than 4 mm, whereas among lesions of 80-100 mm they
disagree by 15 mm and more. The two horizontal lines are too wide on the
left and too narrow on the right. The decision table says the same thing
in numbers. The overall coverage looks fine (`coverage` ≈ 0.95), because a
band that is too wide at one end and too narrow at the other cancels the
two errors in the total. The coverage by thirds of the range exposes it:

```{r constant-cov}
f_id$gof[c("hetero_rho", "hetero_p", "coverage", "cov_low", "cov_mid", "cov_high")]
```

Nearly every difference among small lesions is inside the limits and
roughly one in eight among large lesions is outside.

### Why this matters

Limits of agreement are used to decide whether an observed difference is
more than measurement error. Under RECIST, progression is declared when
the diameter increases by 20% or more. Consider three lesions re-read by
another radiologist:

```{r cases-const}
cases <- data.frame(baseline = c(12, 30, 100), followup = c(20, 37, 114))
cases$change_mm  <- cases$followup - cases$baseline
cases$change_pct <- round(100 * cases$change_mm / cases$baseline)
m_case <- (cases$baseline + cases$followup) / 2
cases$const_limit <- round(loa_band(f_id, mean = m_case)$upper, 1)
cases$verdict_constant <- ifelse(cases$change_mm > cases$const_limit, "real change", "within error")
cases
```

With constant limits, lesion A (12 mm read as 20 mm, a 67% increase and
therefore progression by RECIST) is dismissed as reader disagreement,
while lesion C (100 mm read as 114 mm, a 14% increase, below the
progression threshold) is called a real change. Both verdicts are wrong,
and they are wrong in opposite directions.

## 3. The remedy: choose the scale on evidence

`loa_compare()` fits the conventional limits on each candidate scale,
back-transforms them to the original scale, and tabulates the classical
diagnostics for every candidate.

```{r compare}
cmp <- loa_compare(p_inter)
cmp$gof
```

Candidates are ordered by the heteroscedasticity correlation `hetero_rho`,
the Bland-Altman check of whether the scatter of the differences depends
on the mean. On the identity scale it is strongly positive; on the
logarithmic scale it is strongly negative (the log over-corrects); on the
square-root scale it is close to zero, and the coverage by thirds is even.
The overlay plot shows all four bands on the same scatter:

```{r overlay, fig.cap = "Back-transformed 95% limits of the four candidate transformations."}
plot(cmp, xlab = "Mean of two readers (mm)", ylab = "Difference between readers (mm)")
```

The square-root band follows the fan-shaped spread of the differences.
It is the scale on which the conventional Bland-Altman model holds, so
the conventional limits computed there, and mapped back, are the ones to
use.

## 4. Definitive limits from the variance components

The pairs above are exploratory: pairs formed from the same lesion are
correlated. The definitive estimates come from the hierarchical model
fitted on the chosen scale by `loa_components()`, which separates reader,
lesion-by-reader and residual variance and reports intra- and inter-reader
limits. As in the original study, 5% of observations with the largest
standardized residuals are trimmed.

```{r components}
vc <- loa_components(recist_readings$diameter, recist_readings$lesion,
                     recist_readings$reader, transform = "sqrt", trim = 0.05)
vc
```

`loa_band()` evaluates the curved limits at any lesion sizes of interest:

```{r bands}
sizes <- c(10, 20, 30, 60, 100, 150)
data.frame(size_mm = sizes,
           intra_95 = round(loa_band(vc$intra, mean = sizes)$upper, 1),
           inter_95 = round(loa_band(vc$inter, mean = sizes)$upper, 1),
           constant_inter = round(loa_band(f_id, mean = sizes)$upper, 1))
```

A second reader may plausibly differ by about 4-5 mm on a 10-mm lesion
and by about 14-15 mm on a 100-mm lesion. The constant limit of 8 mm
overstates the error at the small end and understates it at the large
end.

## 5. What changes in the decisions

Re-assess the three lesions with the curved inter-reader limits:

```{r cases-curved}
cases$curved_limit <- round(loa_band(vc$inter, mean = m_case)$upper, 1)
cases$verdict_curved <- ifelse(cases$change_mm > cases$curved_limit, "real change", "within error")
cases$RECIST <- ifelse(cases$change_pct >= 20, "PD (>= 20%)", "SD (< 20%)")
cases[, c("baseline", "followup", "change_mm", "change_pct", "RECIST",
          "const_limit", "verdict_constant", "curved_limit", "verdict_curved")]
```

Lesion A is now recognized as a real change and lesion C as reader
disagreement. Lesion B, near the size where the curved band crosses the
constant limit, is judged the same way by both.

The same comparison over the whole size range shows where a 20% increase
can be produced by reader disagreement alone:

```{r decision-plot, fig.cap = "Minimum change distinguishable from inter-reader error, constant versus curved limits, against the RECIST progression threshold."}
grid <- seq(8, 160, by = 1)
curved <- loa_band(vc$inter, mean = grid)$upper
const  <- loa_band(f_id, mean = grid)$upper
plot(grid, curved, type = "l", lwd = 2, col = "firebrick", ylim = c(0, 40),
     xlab = "Baseline lesion diameter (mm)", ylab = "Increase at re-reading (mm)")
lines(grid, const, lwd = 2, lty = 3, col = "grey30")
lines(grid, 0.2 * grid, lwd = 2, lty = 2, col = "steelblue")
points(cases$baseline, cases$change_mm, pch = 21, bg = "gold", cex = 1.3)
text(cases$baseline, cases$change_mm, c("A", "B", "C"), pos = 3, font = 2)
legend("topleft", bty = "n", lwd = 2, lty = c(1, 3, 2),
       col = c("firebrick", "grey30", "steelblue"),
       legend = c("Inter-reader 95% LOA, square-root band",
                  "Inter-reader 95% LOA, constant",
                  "RECIST progression threshold (+20%)"))
range(grid[0.2 * grid <= const]);  range(grid[0.2 * grid <= curved])
```

Under constant limits a 20% increase is within reader error for lesions up
to about 40 mm; under the curved limits, up to about 50 mm. The two bands
also give different answers to the interchangeability question. If the
acceptable difference between readers is set at 15% of the measurement,
the curved band lies within that tolerance only for lesions above about
90 mm, whereas the constant limit would certify interchangeability from
about 55 mm upwards, over a range where the true error exceeds the
tolerance.

```{r tolerance}
tol <- 0.15 * grid
c(curved_ok_from = min(grid[curved <= tol]), constant_ok_from = min(grid[const <= tol]))
```

## 6. Summary

On data of this kind the conventional limits are not wrong on average,
they are wrong at every particular size, and the decisions built on them
inherit the error. The workflow in **flexloa** replaces the visual choice
of a transformation by a decision table, keeps the conventional
Bland-Altman calculation intact on the chosen scale, and returns the
limits on the original scale where the clinical comparison is made.

```{r summary-workflow, eval = FALSE}
p   <- loa_pairs(diameter, lesion, reader, session, type = "inter")
cmp <- loa_compare(p)                 # decision table + overlay plot
vc  <- loa_components(diameter, lesion, reader, transform = cmp$best)
loa_band(vc$inter, mean = c(10, 30, 100))
```
